Variant calling with transformers

### 2022-06-04 manuscript/manuscript.tex

\subsection{Loss function}

We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare each prediction at window index (position) $i$ to the true base at index $i$. A complexity arises due to ambiguity regarding which haplotype should contain a variant. The two haplotypes present in a sample have no inherent order. In training data variants are assigned arbitrarily to haplotypes, and this randomness in haplotype assignment means the model cannot learn to predict which haplotype should contain a given variant. To combat this ambiguity, we calculate loss in both possible configurations - predicted haplotype 0 with training haplotype 0 and 1 with 1, as well as predicted haplotype 0 with training haplotype 1 and 1 with 0. We select the configuration with the lowest loss and backpropogate gradients only for that configuration.

>>>>

We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare the prediction at window index (position) $i$ to the true base at index $i$, where $i \in 1..g$.  While we obtain satisfactory results with this approach, two issues were apparent. First, predicted haplotype sequences may have inserted or deleted bases relative to the true haplotypes. Because incorrect insertions and deletions cause mismatches at every downstream (higher index) position, such mistakes are penalized more heavily when they occur at lower indexes than higher indexes. Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuracy at high indexes.
